Presheaves, Representables and\\the Yoneda Lemma

### 2017-07-05 slides/slides-10/notes-10.tex
Yet more changes.

    We still have to check naturality, i.e. that for any $B'\xrightarrow{g}B$ the following commutes:
    \begin{align*}
    \xymatrix{
    F_A(B)=\cat A(B,A) \ar[r]^{H_A(g)=-\of g} \ar[d]^{\widetilde a_B} & F_A(B')=\cat A(B',A) \ar[d]_{\widetilde a_B'} \\
    F(B) \ar[r]^{F(g)} & F(B').
                         } 
  \end{align*}
  Since this square is in $\Set$, it suffices to check $(\widetilde a_B'\of H_A(g))(f)=(F(g)\of\widetilde a_B)(f)$ for every $f\in F_A(B)=\cat A(B,A)$. Indeed, this holds:
  \begin{align*}
    (\widetilde a_B'\of H_A(g))(f)&=(F(f\of g))(a)\\
    &=(F(f)\of F(g))(a)&F \text{ functor}\\
    &=F(f)(F(g)(a))\\
    &=(F(g)\of\widetilde a_B)(f).
  \end{align*}
\item Let $a\in F(A)$. Clearly, we get $\widehat{\widetilde a}=a$:
  \begin{align*}
    \widehat{\widetilde a}=\widetilde a_A(1_A)=F(1_A)(a)=1_{F(A)}(a)=a.
  \end{align*}
\item[(d)-(f)] The remaining proof obligations are quite straight-forward to show and are thus left as an exercise.
  \end{enumerate}
\end{proof}

Now, it is easy to show that $H_-$ is full and faithful by instantiating $F$ with a covariant representable functor.
\begin{theorem}
  The Yoneda embedding $H_-\from \cat A\to[\cat A^\op,\Set]$ is full and faithful.
\end{theorem}
\begin{proof}
  We have 
  \begin{align*}
    \widetilde H_f=(H_f)_A(1_A)=f\of 1_A=f.
  \end{align*}
  Hence,
  \begin{align*}
    \cat A(A,A')&\to[\cat A^\op,\Set](H_A,H_A')\\
    f&\mapsto H_f
  \end{align*}
  and
  \begin{align*}
    (\widetilde{-})\from F(A)\to[\cat A^\op,\Set](H_A,F)
  \end{align*}
  are the same function, which are bijective due to the proof of the Yoneda lemma (instantiating $F$ with $H_{A'}$).
\end{proof}

@@

  Then $f$ is an isomorphism in $\cat A$ if and only if $F(f)$ is an isomorphism in $\cat B$.
\end{lemma}
Thus, we immediately get:
\begin{corollary}[Yoneda Principle]
  Let $\cat A$ be locally small and $A,A'\in\cat A$. Then
  \begin{align*}
    H_A\iso H_{A'} \iff A\iso A'
  \end{align*}
\end{corollary}
Note that the direction from right to left is trivial. Thus, the important consequence of this result is that
\begin{align*}
  H_A(B)=\cat A(B,A)\iso \cat  A(B,A')=H_A'(B) \text{ naturally in } B\in\cat A
\end{align*}
implies $A\iso A'$. Consequently, the following slogan is justified:
\begin{quote}
  \textit{``Two objects are the same if and only if they look the same from all viewpoints.''}
\end{quote}
Note that it is really necessary for the isomorphism to be natural in $B\in\cat A$. As the following example illustrates:
TODO
\section{Consequences of the Yoneda Lemma/Principle}
Next, we use the insights just gained and we consider (further) consequences of the Yoneda lemma/principle. In particular, we are going to proof the facts mentioned in the introduction.

@@

\begin{lemma}
  \label{lem:lim}
  For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
  \begin{align*}
    \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
  \end{align*}
  for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
\end{lemma}
\begin{proof}
  First of all, we have to define the shape of the limit, i.e. a small category $\mathbf I$. We take the category of elements of $F$, $F(\cat A)$ with
  \begin{itemize}
  \item objects being pairs $(a,A)$ such that $A\in\cat A$ and $a\in F(A)$;
  \item arrows $f\from(a,A)\to(b,B)$ being arrows $f\from A\to B$ in $\cat A$ such that $F(f)(a)=b$.
  \end{itemize}
  TODO: figure

@@

  To define $H_{\pi(a,A)}=H_A\to F$ for any $(a,A)\in F(\cat A)$ we recall that $[\cat A^\op,\Set](H_A,F)\iso F(A)$, where the isomorphism from the right to the left is given by $(\widetilde -)$ (as defined in the proof of the Yoneda lemma). Thus, we define the $(a,A)$-component of the cocone to be $\widetilde a$. For this to be a cocone we still have to verify that for each $(b,B)\xrightarrow{f}(a,A)$ (i.e. $B\xrightarrow{f}A$ such that $F(f)(b)=a$)
  \begin{align*}
    \xymatrix{
      H_{\pi(b,B)}=H_B \ar[r]^{\widetilde b} \ar[d]_{H_{\pi(f)}=H_f} & F \\
      H_{\pi(a,A)}=H_A \ar[ru]_{\widetilde a} & 
                         } 
  \end{align*}
  commutes. However, by the naturality statement for $A\in\cat A$ in the Yondea lemma we have that 
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{-\of H_f} & [\cat A^\op,\Set](H_B,F) \\
    F(A) \ar[r]^{F(f)} \ar[u]^{(\widetilde-)}& F(B) \ar[u]_{(\widetilde -)}
                         } 
  \end{align*}
  commutes. In particular for any $a\in F(A)$ we have
  \begin{align*}
    \widetilde a\of H_f=\widetilde{F(f)(a)}=\widetilde b
  \end{align*}
  by definition of arrows in $F(\cat A)$.

>>>>

    We still have to check naturality, i.e. that for any $B'\xrightarrow{g}B$ the following commutes:
    \begin{align*}
    \xymatrix{
    F_A(B)=\cat A(B,A) \ar[r]^{H_A(g)=-\of g} \ar[d]^{\widetilde a_B} & F_A(B')=\cat A(B',A) \ar[d]_{\widetilde a_B'} \\
    F(B) \ar[r]^{F(g)} & F(B').
                         } 
  \end{align*}
  Since this square is in $\Set$, it suffices to check $(\widetilde a_B'\of H_A(g))(f)=(F(g)\of\widetilde a_B)(f)$ for every $f\in F_A(B)=\cat A(B,A)$. Indeed, this holds:
  \begin{align*}
    (\widetilde a_B'\of H_A(g))(f)&=(F(f\of g))(a)\\
    &=(F(g)\of F(f))(a)&F \text{ contravariant functor}\\
    &=F(g)(F(f)(a))\\
    &=(F(g)\of\widetilde a_B)(f).
  \end{align*}
\item Let $a\in F(A)$. Clearly, we get $\widehat{\widetilde a}=a$:
  \begin{align*}
    \widehat{\widetilde a}=\widetilde a_A(1_A)=F(1_A)(a)=1_{F(A)}(a)=a.
  \end{align*}
\item[(d)-(f)] The remaining proof obligations are quite straight-forward to show and are thus left as an exercise.
  \end{enumerate}
\end{proof}

Now, it is easy to show that $H_-$ is full and faithful by instantiating $F$ with a covariant representable functor.
\begin{theorem}
  The Yoneda embedding $H_-\from \cat A\to[\cat A^\op,\Set]$ is full and faithful.
\end{theorem}
\begin{proof}
  We have 
  \begin{align*}
    \widehat H_f=(H_f)_A(1_A)=f\of 1_A=f.
  \end{align*}
  Hence,
  \begin{align*}
    \cat A(A,A')&\to[\cat A^\op,\Set](H_A,H_A')\\
    f&\mapsto H_f
  \end{align*}
  and
  \begin{align*}
    (\widetilde{-})\from F(A)\to[\cat A^\op,\Set](H_A,F)
  \end{align*}
  are the same function, which are bijective due to the proof of the Yoneda lemma (instantiating $F$ with $H_{A'}$).
\end{proof}

@@

  Then $f$ is an isomorphism in $\cat A$ if and only if $F(f)$ is an isomorphism in $\cat B$.
\end{lemma}
Thus, we immediately get:
\begin{corollary}[Yoneda Principle]
  Let $\cat A$ be locally small and $A,A'\in\cat A$. Then
  \begin{align*}
    H_A\iso H_{A'} \iff A\iso A'
  \end{align*}
\end{corollary}
Note that the direction from right to left is trivial. Thus, the important consequence of this result is that
\begin{align*}
  H_A(B)=\cat A(B,A)\iso \cat  A(B,A')=H_A'(B) \text{ naturally in } B\in\cat A
\end{align*}
implies $A\iso A'$. Consequently, the following slogan is justified:
\begin{quote}
  \textit{``Two objects are the same if and only if they look the same from all viewpoints.''}
\end{quote}
Note that it is really necessary for the isomorphism to be natural in $B\in\cat A$. As the following example illustrates:
TODO
\section{Consequences of the Yoneda Principle}
Next, we use the insights just gained and we consider (further) consequences of the Yoneda lemma/principle. In particular, we are going to proof the facts mentioned in the introduction.

@@

\begin{lemma}
  \label{lem:lim}
  For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
  \begin{align*}
    \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
  \end{align*}
  for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
\end{lemma}
\begin{proof}
  First of all, we have to define the shape of the limit, i.e. a small category $\mathbf I$. We take the category of elements of $F$, $F(\cat A)$ with
  \begin{itemize}
  \item objects being pairs $(a,A)$ such that $A\in\cat A$ and $a\in F(A)$;
  \item arrows $f\from(a,A)\to(b,B)$ being arrows $f\from A\to B$ in $\cat A$ such that $F(f)(b)=a$.
  \end{itemize}
  TODO: figure

@@

  To define $H_{\pi(a,A)}=H_A\to F$ for any $(a,A)\in F(\cat A)$ we recall that $[\cat A^\op,\Set](H_A,F)\iso F(A)$, where the isomorphism from the right to the left is given by $(\widetilde -)$ (as defined in the proof of the Yoneda lemma). Thus, we define the $(a,A)$-component of the cocone to be $\widetilde a$. For this to be a cocone we still have to verify that for each $(b,B)\xrightarrow{f}(a,A)$ (i.e. $B\xrightarrow{f}A$ such that $F(f)(a)=b$)
  \begin{align*}
    \xymatrix{
      H_{\pi(b,B)}=H_B \ar[r]^{\widetilde b} \ar[d]_{H_{\pi(f)}=H_f} & F \\
      H_{\pi(a,A)}=H_A \ar[ru]_{\widetilde a} & 
                         } 
  \end{align*}
  commutes. However, by the naturality statement for $A\in\cat A$ in the Yondea lemma we have that 
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{-\of H_f} & [\cat A^\op,\Set](H_B,F) \\
    F(A) \ar[r]^{F(f)} \ar[u]^{(\widetilde-)}& F(B) \ar[u]_{(\widetilde -)}
                         } 
  \end{align*}
  commutes. In particular for any $a\in F(A)$ we have
  \begin{align*}
    \widetilde a\of H_f=\widetilde{F(f)(a)}=\widetilde b
  \end{align*}
  by definition of arrows in $F(\cat A)$.
